---
title: "Session 9: Tutorial Differentiated Demand"
author: "Dennis Rickert, MINES Paris-PSL University"
date: "2023-07-17"
output: 
  ioslides_presentation: 
#  html_document: 
#    number_sections: yes
#    toc: yes
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Overview of QEC9X01

-   Session 1: Oligopolistic Competition Refresher
-   Session 2: Oligopolistic Competition Refresher (cont.)
-   Session 3: Introduction into R and Econometrics
-   Session 4: Introduction into R and Econometrics (cont.)
-   Session 5: Estimation of Demand for Homogenous Goods
-   Session 6: Identification of Conduct and Collusion
-   Session 7: Econometrics Tutorial: Conduct Estimation
-   Session 8: Product Differentiation, Multinomial Logit
-   ***Session 9: Differentiated Demand Workshop***
-   Session 10: Market Structure and Entry
-   Session 11: Merger analysis
-   Session 12: Merger analysis - Empirical tutorial in R

## Learning objectives of session 9

### 1. Motivating example of Bertrand-Nash (workhorse)
- Understand theory similar to model of Cournot Conduct
- See applications of the model
- Discuss relevance of good demand estimation

### 2. Heterogeneous demand and differentiated products 
- Understand problems with models of homogenous demand  
- Implement logit and nested logit demand models
- Derive markups from logit demand + Bertrand competition


## How do we proceed in this session?

### ***Bertrand-Nash pricing with differentiated products***
- observe data of prices, market shares, and ownership
- derive costs as function of ownership, market share, elasticity
- obtain markups by subtracting cost from price 
- need to estimate elasticity through demand function

### ***How do we estimate demand?***
- simple case: homogeneous demand (homogeneous products)
- difficult: heterogeneous demand + differentiated products 
  - consumer differ in tastes = variation of elasticities 
  - and firms introduce products depending on taste
  - products have different consumers with different tastes 


## 
**Consider multi-product Bertrand where firms solve** (skip): 
$$
\begin{align*}
\arg \max_{p \in \mathcal{J}_f} \pi_f (\mathbf{p}) &  = \sum_{j \in \mathcal{J}_f} (p_j - c_j) \cdot q_j(\mathbf{p}) \\
 0 & = q_j(\mathbf{p}) + \sum_{k \in \mathcal{J}_f} (p_k - c_k) \frac{\partial q_{k}}{\partial p_j}(\mathbf{p}) \tag{1} \\
\rightarrow BR: p_j(p_{-j}) & =  c_{j} + q_{j}(\mathbf{p}) \left[-\frac{\partial q_{j}}{\partial p_{j}}(\mathbf{p})\right]^{-1} \\
& + \sum_{k \in \mathcal{J}_{f} \setminus j} \left(p_{k}-c_{k}\right) \underbrace{\frac{\partial q_{k}}{\partial p_{j}}(\mathbf{p})\left[-\frac{\partial q_{j}}{\partial p_{j}}(\mathbf{p})\right]^{-1}}_{D_{jk}(\mathbf{p})} \\
\end{align*}
$$

We call $D_{jk}(\mathbf{p}) = \frac{\frac{\partial q_{k}}{\partial p_j}(\mathbf{p})}{\left| \frac{\partial q_{j}}{\partial p_j}(\mathbf{p}) \right|}$ the diversion ratio; $D_{jk} \in [0,1]$.

## 

We can also re-write the best-response in the Lerner Index form:

$$p_j(p_{-j}) = \underbrace{\frac{1}{1+1/\epsilon_{jj}(\mathbf{p})}}_{\text{Markup}} \left[ c_j + \underbrace{\sum_{k \in \mathcal{J}_{f} \setminus j}  (p_k-c_k) \cdot  D_{jk} (\mathbf{p})}_{Opportunity~Costs} \right]$$

- we have usual Lerner markup given by inverse elasticity
- multi-product firm has additional market power from opportunity costs of raising the price $p_j$, 
  - possible if clients switch to other products $k$ of same firm 
  - $D_{jk}$ gives proportion of clients that stay within firm
  - for which firm gets the markup $p_k-c_k$
  
## Differentiated Products Bertrand

We take the FOC w.r.t. price for all $j$ and stack them up
$$
\begin{gather}
	\begin{pmatrix} 
		q_{j=1} \\
		 \vdots \\
		q_{j=J} 
	\end{pmatrix}
	+
	\begin{pmatrix} 
		\frac{\partial q_{j=1}}{\partial p_{j=1}} & ... & \frac{\partial q_{j=J}}{\partial p_{j=1}} \\
		 \vdots & \ddots & \vdots \\
		\frac{\partial q_{j=1}}{\partial p_{j=J}} & ... & \frac{\partial q_{j=J}}{\partial p_{j=J}} 
	\end{pmatrix}
	\begin{pmatrix} 
		p_{j=1}-c_{j=1} \\
		 \vdots \\
		p_{j=J}-c_{j=J}
	\end{pmatrix}
	=0	
\end{gather}
$$
...where the above example corresponds to a case of monopoly

- all products belong to the same firm
- demand lost on product moves to product of the same firm
- internalizing this "demand loss" leads to higher prices


## 
which we can solve for the price cost margin: 
$$
\begin{gather}
	\begin{pmatrix} 
		p_{j=1}-c_{j=1} \\
		 \vdots \\
		p_{j=J}-c_{j=J}
	\end{pmatrix}
= -   
	\begin{pmatrix} 
		\frac{\partial q_{j=1}}{\partial p_{j=1}} & ... & \frac{\partial q_{j=J}}{\partial p_{j=1}} \\
		 \vdots & \ddots & \vdots \\
		\frac{\partial q_{j=1}}{\partial p_{j=J}} & ... & \frac{\partial q_{j=J}}{\partial p_{j=J}} 
	\end{pmatrix}
\end{gather}^{-1}
	\begin{pmatrix} 
		q_{j=1} \\
		 \vdots \\
		q_{j=J} 
	\end{pmatrix}
$$
<br> 

... which is related to the Lerner Index

- we can also solve it for marginal costs

## Differentiated Products Bertrand

It is helpful to define the matrix \( \Delta \) with entries of demand derivatives (not elasticities!):

\[
\Delta_{(j,k)}(\mathbf{p}) = \left\{\begin{array}{lr}
         - \frac{\partial q_{j}}{\partial p_k}(\mathbf{p}) & \text{for }  (j,k) \in \mathcal{J}_f\\
       	  \quad 0 & \text{for } (j,k) \notin \mathcal{J}_f
        \end{array} \right\}
\]

- $\Delta$ is a full entry matrix full in case of monopoly/full collusion  
- $\Delta$ is a diagonal matrix in case of single-product firms 

<br>
We can re-write the FOC in matrix form:

\[
(\mathbf{p}-\mathbf{c}) = - \Delta(\mathbf{p})\cdot q(\mathbf{p}) 
\]


## Ownership Matrix Example

Assume we have 3 products controlled by 2 firms. We can represent the ownership matrix \( \Delta^B \) as follows:

$$
\Delta^B = \left\{
  \begin{matrix}
    - \frac{{\partial q_1}}{{\partial p_1}} &  \frac{{\partial q_1}}{{\partial p_2}} & 0 \\
     \frac{{\partial q_2}}{{\partial p_1}} & - \frac{{\partial q_2}}{{\partial p_2}} & 0 \\
    0 & 0 & - \frac{{\partial q_3}}{{\partial p_3}}
  \end{matrix}
\right\}
$$

- Products 1 and 2 are controlled by firm 1, so we have non-zero entries in the first two rows and first two columns.
- Product 3 is controlled by firm 2, so we have non-zero entry in the last row and last column.
- The zeros represent the fact that there's no ownership relationship for the corresponding product pairs.


## What to learn from demand systems?

In monopoly this matrix becomes:
$$
\Delta^M = \left\{
  \begin{matrix}
    - \frac{{\partial q_1}}{{\partial p_1}} &  \frac{{\partial q_1}}{{\partial p_2}} &  \frac{{\partial q_1}}{{\partial p_3}} \\
     \frac{{\partial q_2}}{{\partial p_1}} & - \frac{{\partial q_2}}{{\partial p_2}} &  \frac{{\partial q_3}}{{\partial p_3}} \\
    \frac{{\partial q_1}}{{\partial p_3}} &  \frac{{\partial q_2}}{{\partial p_3}} & - \frac{{\partial q_3}}{{\partial p_3}}
  \end{matrix}
\right\}
$$
different $p-c$ and $c$ for monopoly and competition: 

$$(\mathbf{p^*}-\mathbf{c^B}) = \mathbf{\Delta^B (p)}^{-1} \mathbf{q^*(p)}$$

$$(\mathbf{p^*}-\mathbf{c^M}) = \mathbf{\Delta^M (p)}^{-1} \mathbf{q^*(p)}$$

... at observed values $p^*$ and $q^*$. 

## Applications
Multiproduct Demand system estimation is probably the most important contribution from the New Empirical IO literature:

- Welfare analysis, pass-on of taxation, value of advertising, price effects of mergers, and many more applications

<br>

Last 5-10 years has seen successful export to other fields: 

- Marketing, Labor. Trade, Healthcare, Education, 

<br>

Anywhere consumers face a number of options and ``prices'':

- doctors, hospitals, schools, mutual funds, etc.


## Data Sources
**Aggregate data** (market level data on \( (P_j,Q_j) \))

- Many supermarket "scanner" datasets: Nielsen, IRI, ... 

Or **micro data** panel of data with same individuals over time.

  - Best example is Nielsen Homescan Consumer Panel data.
  - Visa/Mastercard datasets
  - Medicare

Or **aggregate purchase data** plus **micro data** from a survey 

- ie: asking people who purchased GM cars which other cars they were considering.
- Using scanner data for alcohol purchases and comparing to consumption surveys by income and education.


## Taxonomy of Demand Systems

Lots of different ways to classify demand systems 

- We saw homogeneous demand + homogeneous products 
- Focus here: heterogeneous demand + differentiated products 
  - product characteristics space

## Homogenous Demand

### Definition

- All consumers have the same preferences
- Price elasticity is constant across consumers

### Modeling Approach

- Linear demand model
- log-lin
- log log

Examples were given in last lecture 

- we had just one parameter and one elasticity for all products

## Heterogenous Demand

### Definition

- Consumers have different preferences and elasticities
- Elasticity varies with price and/or consumer characteristics

### Modeling approach
- Logit Model
- Nested logit model
- Random-coefficient logit model

### Importance of capturing Heterogeneity
- Accurate analysis of demand and market power
- Allows for more accurate policy analysis


# The Logit Model

## Utility Function
For consumer \(i\) choosing product \(j\):
$$
U_{ij} = X_j \beta - \alpha p_j + \xi_j + \epsilon_{ij}
$$
where:
- \(X_j\) is product characteristics
- \(p_j\) is price
- \(\xi_j\) is unobserved product characteristics
- \(\epsilon_{ij}\) is idiosyncratic consumer preference

### Decision Making
Consumer \(i\) chooses product \(j\) if \(U_{ij} > U_{ik}\) for all \(k \neq j\).

## Choice Probabilities & Market Shares
The probability that consumer \(i\) chooses product \(j\) is:
$$
s_{ij} = \frac{\exp(X_j \beta - \alpha p_j + \xi_j)}{\sum_{k=1}^J \exp(X_k \beta - \alpha p_k + \xi_k)}
$$
Market share for product \(j\) is the average across consumers:
$$
s_j = \int s_{ij} di
$$

## Estimation Equation
$$
\log(s_j) - \log(s_0) = X_j \beta - \alpha p_j + \xi_j
$$
where \(s_0\) is the outside option's market share.

## Elasticities
Own-price elasticity:
$$
\eta_{jj} = \alpha \cdot p_j \cdot (1 - s_j)
$$
Cross-price elasticity:
$$
\eta_{jk} = \alpha \cdot p_k \cdot s_k \quad \text{for } j \neq k
$$


## Disadvantage of Logit
The primary disadvantage of the standard Logit model compared to the Nested Logit model is its restrictive substitution patterns. This drawback is termed the Independence of Irrelevant Alternatives (IIA) property.

See examples in class (red bus / blue bus)

- introducing a blue bus, consumers will more likely switch from red bus than from bike 

## Nested Logit as a Solution

**Relaxing IIA**: Allows correlated error terms within groups of similar alternatives ("nests").

- Nested Logit: More flexible substitution patterns but needs correct nesting structures
- E.g., we can define a nest bus vs bike



# The Nested Logit Model

## Utility Function
The Nested Logit divides products into \(G\) groups. For consumer \(i\) choosing product \(j\) in group \(g\):
$$
U_{ij} = X_j \beta - \alpha p_j + \xi_j + \sigma \cdot \eta_{ig} + \epsilon_{ij}
$$
where \(\eta_{ig}\) captures correlation in utility across products in group \(g\).


## Estimation Equation

For the Nested Logit model, the estimation equation becomes: 
$$
\log(s_j) - \log(s_0) = X_j \beta - \alpha p_j + \xi_j + \sigma \cdot \log s_{j|g(j)}
$$

- where $s_{j|g(j)}$ is the share of product $j$ in group $g$ 
- which reflects the increased utility from products within the same group \(g\).

<br>

the nesting parameter is endogenous as it depends on price

- we need an additional instrument


## Elasticities
***Main takeaway***: Nested Logit allows for different cross-price elasticities depending on whether products are in same group


1. **Own-price elasticity** for product \(j\) in group \(g\):
$$
\eta_{jj} = \alpha \cdot p_j \cdot \left(1 - s_{j|g} - \sigma \cdot s_{j|g} \cdot s_g\right)
$$

2. **Cross-price elasticity** for products \(j\) and \(k\) both in group \(g\):
$$
\eta_{jk} = \alpha \cdot p_k \cdot s_{k|g} \cdot (s_{j|g} + \sigma \cdot s_{j|g} \cdot s_g) \quad \text{for } j \neq k
$$

3. **Cross-price elasticity** for product \(j\) in group \(g\) and product \(k\) in group \(h\):
$$
\eta_{jk} = \alpha \cdot p_k \cdot s_{k|h} \cdot s_{j|g} \quad \text{for } g \neq h
$$


# Differentiated demand, logit exercise

## 1. Import data

import the data into a data frame, called 'cameras', 

```{r import, echo=TRUE}
#install.packages("flextable")

library(haven)
cameras <- read.csv("cameras.csv")
```

This is a data set of digital cameras per EU country and brand

- prices, shares, characteristics at a single date
- charactersitcis: type, slr, elect, optical, pixeltot

## Tabulating sales by brand and country

```{r tables, echo=TRUE, message=FALSE}
library(dplyr)
library(flextable)

cameras %>%
  group_by(Country) %>%
  summarise(Sales = sum(salesunits),
            "Average price" = mean(priceur)) %>%
  ungroup() %>%
  flextable()
```

## Sales and price by brand 

```{r brand_group, echo=TRUE}
cameras %>%
  group_by(Brand) %>%
  summarise("Sales" = sum(salesunits),
            "Average price" = mean(priceur)) %>%
  ungroup() %>%
  flextable()

```

## Prices of brand per country

```{r brand_country_group, echo=TRUE, message=FALSE}
cameras %>%
  group_by(Brand, Country) %>%
  summarise("Sales" = sum(salesunits),
            "Average price" = mean(priceur)) %>%
  ungroup() %>%
  flextable()
```


## 

Before we proceed to state-of-the art estimation 

- of logit and nested logit, 

<br>

We look at ``reduced-form'' hedonic regressions

- kind of homogeneous demand function
- which can be extended to heterogenous demand for differentiated products at very high costs

## Hedonic price/demand regressions

hedonic: ("pleasure/utility") regressions to uncover correlations, 

- Estimate how much people value different attributes 
- What determines prices and quantity of a product?

<br>

Limitations: 

- no underlying theory which guides choice of variables 
  - no endogeneity, e.g., economic theory tells us about OVB
  - no consumer decision making, no utility maximizing
  - no profit maximization
- throwing in arbitrary variables to "p-hack" effect of interest 


## Hedonic regressions

We define country and brand as factor variables instead of including them manually

```{r iv1, echo=TRUE}
# Convert brand and country into factors:
cameras$brand <- as.factor(cameras$Brand)
cameras$country <- as.factor(cameras$Country)
```

## Hedonic price regressions

We can start by analyzing what is driving the prices

```{r hedonic2_a, echo=TRUE, message=FALSE}
hedonic2 <- lm(lpriceur ~  type + slr + elect + optical + pixeltot
                 , data = cameras)
```

## 
```{r hedonic2_b, echo=TRUE, message=FALSE}
summary(hedonic2)
```

## 

Decomposing the prices into country means

- this is the same as calculating the mean prices per country

```{r hedonic3_a, echo=TRUE, message=FALSE}
hedonic3 <- lm(lpriceur ~ country
               , data = cameras)
```

<br>

don't forget that we have coded country as a factor variable 


##
```{r hedonic3_b, echo=TRUE, message=FALSE}
summary(hedonic3)
```

## 

Now we can define some kind of demand model

```{r hedonic4_a, echo=TRUE, message=FALSE}
hedonic4 <- lm(lsalesunits ~ lpriceur 
               + type + slr + elect + optical + pixeltot + country + brand
               , data = cameras)
```

##
```{r hedonic4_b, echo=TRUE, message=FALSE}
summary(hedonic4)
```

## Limitations

These are reduced-form correlations which do not allow to: 

- adress endogeneity, e.g., OVB of supply side
- model interactions of demand and supply equations
- derive counterfactual analysis for policy advice
- address consumer heterogeneity in tastes
- back out costs and markups consistent with quantity/price

## this as a homogeneous demand model 

... without endogeneity. Only one parameter of elasticity.

- consumers only care about price and all products are the same, then the price of other products doesn't matter, 
  - leading to a cross-price elasticity of zero.

## heterogeneous demand model 

Let's introduce heterogeneity by letting demand depend on prices of other products

```{r, echo=TRUE, message=FALSE}
library(dplyr)

# Step 1: Grouping
cameras_grouped <- group_by(cameras, Country, period)

# Step 2: Mutate
cameras_grouped <- mutate(cameras_grouped, own_price = priceur,
                     cross_price_avg = (sum(priceur) - own_price) / (n() - 1))

# Step 3: Ungroup
cameras_final <- ungroup(cameras_grouped)

```

## 

we include the price of other products into the regression 

```{r, echo=TRUE, message=FALSE}
model <- lm(salesunits ~ own_price + cross_price_avg
            + country + brand 
            + type + slr + elect + optical + pixeltot 
            , data = cameras_grouped)
```

## 

```{r, echo=TRUE,message=FALSE}
summary(model)  
```

## Homogenous demand

This is special version of AIDS (Almost Ideal Demand System)

- we get (homogeneous) own-price effects, 
- for cross price effects, we include price of other products 
- we use one additional price variable per product

<br>

easy-peasy, right?


## BUT 

```{r, echo=TRUE}
num_distinct <- length(unique(cameras$brand))
print(num_distinct)

num_distinct <- length(unique(cameras$model))
print(num_distinct)
```

- introduce for each product $j$ in a country $1+7$ or $1+807$ new coefficients. Full substitution: $7^2$ or $807^2$
- assuming that elasticities are time invariant

## There is hope 

Models of characteristics space: reduce number of parameters

- Instead of estimating separate parameters for each product  
- we estimate utilities or coefficients on product characteristics. 

<br> 

For example, if you have 1,000 products, 

- but only 10 meaningful characteristics that define products
- then just assume that those parameters define preferences 
- and estimate those 10 parameters

This is what logit, nested logit, and BLP do

## 

Remember the formula: need to calculate $log(s_j)-log(s_0)$, 

- where $log(s_0)$ is the outside good, often market size
- if you observe a sample of 10 million clients in a country, then the outside good (probability of not buying) could be the population size
- here we take it as given


```{r logit1_reg_a, echo=TRUE, message=FALSE}
logit1 <- lm(ls ~  priceur 
             + type + slr + elect + optical + pixeltot 
             + country + brand
             , data = cameras)
```

## 
```{r logit1_reg_b, echo=TRUE}
summary(logit1)
```

## 

```{r, echo=TRUE, message=FALSE}
library(dplyr)

# Calculate sum of characteristics in other countries for each country
result <- cameras %>%
  group_by(Country) %>%
  mutate(sum_optical_other = sum(cameras$optical[cameras$Country != first(Country)]),
  sum_elect_other = sum(cameras$elect[cameras$Country != first(Country)]),        
  sum_type_other = sum(cameras$type[cameras$Country != first(Country)]),
  sum_slr_other = sum(cameras$slr[cameras$Country != first(Country)]),
  sum_pixeltot_other = sum(cameras$pixeltot[cameras$Country != first(Country)])) %>%
  ungroup()
```


## 
```{r iv2, echo=TRUE, message=FALSE}
library(AER)

ivlogit1 <- ivreg(ls ~  
                    + priceur + type + slr + elect+ optical + pixeltot + country | 
                  country + type + slr + elect + optical + pixeltot  
                  + sum_optical_other + sum_pixeltot_other + sum_elect_other + sum_type_other + sum_slr_other
                , data = result) 

```



## 
```{r, echo=TRUE}
summary(ivlogit1)
```


## Nested logit
```{r nlogit_reg_a, echo= TRUE}
logit2 <- lm(ls ~  + priceur + lsj_g 
             + type + slr + elect+optical + pixeltot 
             + brand + country , data = result)
```

## 
```{r nlogit_reg_b, echo= TRUE}
summary(logit2)
```

## 
```{r nlogit3_a, echo=TRUE, message=FALSE}
ivlogit2 <- ivreg(ls ~ priceur + lsj_g 
                  + type + slr + elect + optical + pixeltot 
                  + brand + country | 
                    lsj_g + type + slr + elect + optical + pixeltot + brand + country
                  + sum_optical_other + sum_pixeltot_other + sum_elect_other + sum_type_other + sum_slr_other 
                , data = result) 
```
## 
```{r nlogit3_b, echo=TRUE, message=FALSE}
summary(ivlogit2)
```

## Price elasticities based on logit

```{r elasticities, echo=TRUE, message=FALSE}
library("broom")

sum <- tidy(ivlogit1)

alpha = coef(summary(ivlogit1))["priceur","Estimate"]
result$own = alpha*result$priceur*(1-result$share)
result$cross = -alpha*result$priceur*result$share

```

##
```{r neat, echo=TRUE, message=FALSE}
result %>%
  group_by(Country, Brand) %>%
  summarise(mean_own = mean(own)) %>%
  tidyr::pivot_wider(id_cols= Country, names_from = Brand, values_from = mean_own) %>%
  ungroup() %>%
  flextable()

```


```{r elasticities_cross, echo=TRUE}

result %>%
  group_by(Country, Brand) %>%
  summarise(mean_cross = mean(cross)) %>%
  tidyr::pivot_wider(id_cols= Country, names_from = Brand, values_from = mean_cross) %>%
  ungroup() %>%
  flextable()


```

## Matrix of Market Share Derivatives 

... (also called the Jacobian):

```{r, echo=TRUE}
cameras_austria <- result %>% filter(country == "AUSTRIA")

s <- cameras_austria$share
diagM <- alpha * s * (1 - s)
offdiagM <- -alpha * outer(s, s)
Jacobian <- matrix(0, length(s), length(s))
diag(Jacobian) <- diagM
Jacobian[Jacobian == 0] <- offdiagM[Jacobian == 0]
```

## Ownership Matrix

```{r, echo=TRUE}
brand <- cameras_austria$Brand
Ownership <- outer(brand, brand, FUN = "==") * 1
```

## Markups and costs

```{r, echo=TRUE}
p <- cameras_austria$priceur
mu <- -solve(Jacobian*Ownership) %*% s
c <- p - mu

print(mean(c))
print(mean(mu))

lerner=(p-c)/p
print(mean(lerner))
```
## 
```{r, echo=TRUE}
# merger to full monopoly or full collusion

mu_m <- -solve(Jacobian) %*% s
c_m <- p - mu_m

print(mean(c_m))
print(mean(mu_m))

lerner_m=(p-c_m)/p
print(mean(lerner_m))
```

