---
title: "Differentiated demand, logit exercise"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1 Import & clean data, produce descriptive statistics

## Import data

First, import the data into a data frame, called 'cameras_raw', and use the 'dim', 'str' and 'head' functions to describe the dataset:

```{r import}
library(haven)
cameras_raw <- read_dta("differentiated/differentiated.dta")
```

## create population variable

Next, create the population variable, and save a new dataframe, 'cameras':

```{r population, echo=FALSE, message=FALSE}
library(dplyr)
cameras <- cameras_raw %>%
  mutate(population = 0,
         population = case_when(country=="AUSTRIA" ~ 8451900,
                                country=="BELGIUM" ~ 11161600,
                                country=="BULGARIA" ~ 7284600,
                                country=="CZECH REPUBLIC" ~ 10516100,
                                country=="DENMARK" ~ 5602600,
                                country=="FRANCE" ~ 65633200,
                                country=="FINLAND" ~ 5426700,
                                country=="GERMANY" ~ 80523700,
                                country=="GREAT BRITAIN"~ 63730100,
                                country=="GREECE"~ 11062500,
                                country=="HUNGARY"~ 9908800,
                                country=="ITALY"~ 59685200,
                                country=="IRELAND"~ 4591100,
                                country=="NETHERLANDS"~ 16779600,
                                country=="POLAND"~ 38533300,
                                country=="PORTUGAL"~ 10487300,
                                country=="ROMANIA"~ 20057500,
                                country=="SLOVAKIA"~ 5410800,
                                country=="SLOVENIA"~ 2058800,
                                country=="SPAIN"~ 46704300,
                                country=="SWEDEN"~ 9555900),
        population = population/1000) %>%
        rename(Brand = brand,
               Country = country)

```

## Create logs of prices and quantities

Basic plot of tea price against time:

```{r logs, echo=FALSE}
cameras <- cameras %>% 
  mutate(lpriceur=log(priceur),
         lsalesunits=log(salesunits))

```

## Check uniqueness of brand+model

```{r unique, echo=FALSE}
cameras <- cameras %>%
  group_by(model, Brand) %>%
  mutate(product = cur_group_id()) %>%
  ungroup()
  

```

```{r unique_check}
cameras %>%
  group_by(product, Country) %>%
  summarise(count = n(), 
            sum = sum(salesunits), 
            mean = mean(salesunits)) %>%
  ungroup()
```



## Rescaling sales


```{r rescale}

cameras <- cameras %>%
  mutate(salesunits = salesunits / 1000,
    group = elect)

```

## Create variables: total sales, market size and dependent variable for MNL, nested logit


```{r derived_variables, message=FALSE}
country_sales <- cameras %>%
  group_by(Country) %>%
  mutate(salesunits_t = sum(salesunits)) %>%
  ungroup()

cameras <- cameras %>%
  inner_join(country_sales) %>%
  mutate(MSIZE = 0.4 * population,
         share = salesunits / MSIZE,
         share0 = (MSIZE - salesunits_t)/MSIZE,
         ls=log(share/share0))

```

## Nested logit - shares within a segment


```{r nest_shares, message=FALSE}
country_nest <- cameras %>%
  group_by(Country, group) %>%
  mutate(salesunits_g = sum(salesunits)) %>%
  ungroup()

cameras <- cameras %>%
  inner_join(country_nest) %>%
  mutate(lsj_g = log(salesunits / salesunits_g)) %>%
  mutate(con = 1)

```

## 

Export data set

```{r}
write.csv(cameras, "cameras.csv", row.names=FALSE)
```
