---
title: "Differentiated demand, logit exercise"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1 Import & clean data, produce descriptive statistics

## Import data

First, import the data into a data frame, called 'cameras_raw', and use the 'dim', 'str' and 'head' functions to describe the dataset:

```{r import}
library(haven)
cameras_raw <- read_dta("differentiated.dta")

```

## create population variable

Next, create the population variable, and save a new dataframe, 'cameras':

```{r population, echo=FALSE}
library(dplyr)
cameras <- cameras_raw %>%
  mutate(population = 0,
         population = case_when(country=="AUSTRIA" ~ 8451900,
                                country=="BELGIUM" ~ 11161600,
                                country=="BULGARIA" ~ 7284600,
                                country=="CZECH REPUBLIC" ~ 10516100,
                                country=="DENMARK" ~ 5602600,
                                country=="FRANCE" ~ 65633200,
                                country=="FINLAND" ~ 5426700,
                                country=="GERMANY" ~ 80523700,
                                country=="GREAT BRITAIN"~ 63730100,
                                country=="GREECE"~ 11062500,
                                country=="HUNGARY"~ 9908800,
                                country=="ITALY"~ 59685200,
                                country=="IRELAND"~ 4591100,
                                country=="NETHERLANDS"~ 16779600,
                                country=="POLAND"~ 38533300,
                                country=="PORTUGAL"~ 10487300,
                                country=="ROMANIA"~ 20057500,
                                country=="SLOVAKIA"~ 5410800,
                                country=="SLOVENIA"~ 2058800,
                                country=="SPAIN"~ 46704300,
                                country=="SWEDEN"~ 9555900),
        population = population/1000) %>%
        rename(Brand = brand,
               Country = country)

```

## Tabulating sales by brand and country


```{r tables}
library(flextable)
cameras %>%
  group_by(Country) %>%
  summarise(Sales = sum(salesunits),
            "Average price" = mean(priceur)) %>%
  ungroup() %>%
  flextable()
  
```

```{r brand_group}

cameras %>%
  group_by(Brand) %>%
  summarise("Sales" = sum(salesunits),
            "Average price" = mean(priceur)) %>%
  ungroup() %>%
  flextable()

```

```{r brand_country_group}

cameras %>%
  group_by(Brand, Country) %>%
  summarise("Sales" = sum(salesunits),
            "Average price" = mean(priceur)) %>%
  ungroup() %>%
  flextable()

```


## Create country dummies

```{r country, echo=FALSE}

library(sjmisc)
cameras <- cameras %>% 
  to_dummy(Country, suffix = "label") %>% 
  bind_cols(cameras)

```

## Create logs of prices and quantities

Basic plot of tea price against time:

```{r logs, echo=FALSE}
cameras <- cameras %>% 
  mutate(lpriceur=log(priceur),
         lsalesunits=log(salesunits))

```

## Hedonic price regression


```{r hedonic_model}
model1_data <- cameras %>%
  select(lsalesunits, lpriceur, Country_BELGIUM, Country_BULGARIA, `Country_CZECH REPUBLIC`, Country_DENMARK, Country_FINLAND , Country_FRANCE, Country_GERMANY,
         `Country_GREAT BRITAIN` , `Country_GREECE` , Country_HUNGARY ,  Country_IRELAND , Country_ITALY ,  Country_NETHERLANDS , Country_POLAND , Country_PORTUGAL,
         Country_ROMANIA , Country_SLOVAKIA , Country_SLOVENIA , Country_SPAIN , Country_SWEDEN , type , slr , elect , optical , pixeltot)

```

```{r hedonic1}
colnames(cameras)
hedonic1 <- lm(lpriceur ~ . -lsalesunits -type -slr -elect -optical -pixeltot, data = model1_data)
summary(hedonic1)
```

```{r hedonic2}
hedonic2 <- lm(lsalesunits ~ . -type -slr -elect -optical -pixeltot, data = model1_data)
summary(hedonic2)

```


```{r hedonic3}
hedonic3 <- lm(lpriceur ~ . -lsalesunits, data = model1_data)
summary(hedonic3)

```

```{r hedonic4}
hedonic4 <- lm(lsalesunits ~ ., data = model1_data)
summary(hedonic4)

```




## Create a list of characteristics variables

```{r characteristics}
characteristics <- c("variables", "type", "slr", "elect", "optical", "pixeltot")
```


## Check uniqueness of brand+model

```{r unique, echo=FALSE}
cameras <- cameras %>%
  group_by(model, Brand) %>%
  mutate(product = cur_group_id()) %>%
  ungroup()
  

```

```{r unique_check}
cameras %>%
  group_by(product, Country) %>%
  summarise(count = n(), 
            sum = sum(salesunits), 
            mean = mean(salesunits)) %>%
  ungroup()
```



## Rescaling sales


```{r rescale}

cameras <- cameras %>%
  mutate(salesunits = salesunits / 1000,
    group = elect)

```

## Nesting variable - segment

Consider the price of coffee and wages:

```{r nest}


```

## Create variables: total sales, market size and dependent variable for MNL, nested logit


```{r derived_variables}
country_sales <- cameras %>%
  group_by(Country) %>%
  mutate(salesunits_t = sum(salesunits)) %>%
  ungroup()

cameras <- cameras %>%
  inner_join(country_sales) %>%
  mutate(MSIZE = 0.4 * population,
         share = salesunits / MSIZE,
         share0 = (MSIZE - salesunits_t)/MSIZE,
         ls=log(share/share0))

```

## Nested logit - shares within a segment


```{r nest_shares}
country_nest <- cameras %>%
  group_by(Country, group) %>%
  mutate(salesunits_g = sum(salesunits)) %>%
  ungroup()

cameras <- cameras %>%
  inner_join(country_nest) %>%
  mutate(lsj_g = log(salesunits / salesunits_g)) %>%
  mutate(con = 1)

```

## define instrumental variables  


```{r instruments, echo=FALSE, include = FALSE }
country_char <- cameras %>%
  group_by(Country) %>%
  mutate(sum1_con = sum(con),
         sum1_type = sum(type),
         sum1_slr = sum(slr),
         sum1_elect = sum(elect),
         sum1_optical = sum(optical),
         sum1_pixeltot = sum(pixeltot))

country_brand_char <- cameras %>%
  group_by(Country, Brand) %>%
  mutate(sum2_con = sum(con),
         sum2_type = sum(type),
         sum2_slr = sum(slr),
         sum2_elect = sum(elect),
         sum2_optical = sum(optical),
         sum2_pixeltot = sum(pixeltot))


country_group_char <- cameras %>%
  group_by(Country, group) %>%
      mutate(sum3_con = sum(con),
        sum3_type = sum(type),
         sum3_slr = sum(slr),
         sum3_elect = sum(elect),
         sum3_optical = sum(optical),
         sum3_pixeltot = sum(pixeltot))

country_group_brand_char <- cameras %>%
  group_by(Country, group, Brand) %>%
        mutate(sum4_con = sum(con),
               sum4_type = sum(type),
         sum4_slr = sum(slr),
         sum4_elect = sum(elect),
         sum4_optical = sum(optical),
         sum4_pixeltot = sum(pixeltot))


cameras <- cameras %>%
  inner_join(country_char) %>%
  inner_join(country_brand_char) %>%
  inner_join(country_group_char) %>%
  inner_join(country_group_brand_char)

```

## Create brand dummies

```{r brands}
cameras <- cameras %>%
  mutate(i1_type = sum2_type - type,
         i2_type = sum1_type - sum2_type,
         i3_type = sum4_type - type,
	       i4_type = sum3_type - sum4_type,
	       i7_type = type^2
         )

```

```{r for_example}

cameras <- cameras %>%
  relocate(con, type, slr, elect, optical, pixeltot) 
colnames(cameras) # find the cameras characteristics variables

for (var in names(cameras[1:6])) { # loop over all of the variable names
  cameras %>% 
    count(.data[[var]]) %>% print() # '.data' is a pronoun that calls the 'cameras' dataset
}


for (var in names(cameras[1:6])) {
  # create our summation variables dynamically from the variable name
  sumvar1 = paste0("sum1_",var) # paste0 is the same as paste but assumes 'sep = ""'
  sumvar2 = paste0("sum2_",var)
  sumvar3 = paste0("sum3_",var)
  sumvar4 = paste0("sum4_",var)
  # create our new characteristics instrument name dynamically from the variable name
  newvar1 = paste0("i1_",var)
  newvar2 = paste0("i2_",var)
  newvar3 = paste0("i3_",var)
  newvar4 = paste0("i4_",var)
  newvar5 = paste0("i7_",var)
  # create our instruments and assign our dynamic variable name
  cameras <- cameras %>% 
    mutate(!!newvar1 := .data[[sumvar2]] - .data[[var]], # the !! means use the dynamic variable name created above
           !!newvar2 := .data[[sumvar1]] - .data[[sumvar2]], 
           !!newvar3 := .data[[sumvar4]] - .data[[var]],
           !!newvar4 := .data[[sumvar3]] - .data[[sumvar4]],
           !!newvar5 := .data[[var]]^2
    )
}

```

## Create brand dummies

```{r brand_dummy}

cameras <- cameras %>%
  to_dummy(Brand, suffix = "label") %>%
  bind_cols(cameras)

```



## Logit regression

```{r logit_reg}

colnames(cameras)
model2_data <- cameras %>%
  select(c("ls",
           starts_with("Brand_"),
           starts_with("Country_"),
           "priceur",
           "type", "slr", "elect", "optical", "pixeltot"))

logit1 <- lm(ls ~ . -Brand_CANON -Country_AUSTRIA, data = model2_data)
summary(logit1)
  
library("AER")


```

```{r iv1}
model3_data <- cameras %>%
  select(c("ls",
           starts_with("Brand_"), 
           starts_with("Country_"),
           "priceur",
           "type", "slr", "elect", "optical","pixeltot",
           starts_with("i1"),
           starts_with("i2"),
           starts_with("i7"))) 
colnames(model3_data)
```
```{r iv2}

ivlogit1 <- ivreg(ls ~  Brand_FUJIFILM + Brand_KODAK + Brand_NIKON + Brand_OLYMPUS + Brand_PANASONIC + Brand_SAMSUNG + Brand_SONY  
                +Country_BELGIUM + Country_BULGARIA + `Country_CZECH REPUBLIC` + Country_DENMARK + Country_FINLAND + Country_FRANCE + Country_GERMANY 
                + `Country_GREAT BRITAIN` + Country_GREECE + Country_HUNGARY + Country_IRELAND + Country_ITALY + Country_NETHERLANDS + Country_POLAND + Country_PORTUGAL 
                + Country_ROMANIA + Country_SLOVAKIA + Country_SLOVENIA + Country_SPAIN + Country_SWEDEN+ priceur + type + slr + elect+optical + pixeltot + priceur | 
                Brand_FUJIFILM + Brand_KODAK + Brand_NIKON + Brand_OLYMPUS + Brand_PANASONIC + Brand_SAMSUNG + Brand_SONY 
                + Country_BELGIUM + Country_BULGARIA + `Country_CZECH REPUBLIC` + Country_DENMARK + Country_FINLAND + Country_FRANCE + Country_GERMANY 
                + `Country_GREAT BRITAIN` + Country_GREECE + Country_HUNGARY + Country_IRELAND + Country_ITALY + Country_NETHERLANDS + Country_POLAND + Country_PORTUGAL 
                + Country_ROMANIA + Country_SLOVAKIA + Country_SLOVENIA + Country_SPAIN + Country_SWEDEN + type + slr + elect+optical + pixeltot  
                  #instruments start here 
                + i1_con + i2_con + i7_con + i1_type + i2_type + i7_type + i1_slr + i2_slr + i7_slr + i1_elect + i2_elect + i7_elect 
                + i1_optical + i2_optical + i7_optical + i1_pixeltot + i2_pixeltot + i7_pixeltot 
                , data = model3_data) # excluded Brand_CANON + Country_AUSTRIA

summary(ivlogit1)

```



## Nested logit


```{r nlogit_reg, echo= FALSE}

model4_data <- cameras %>%
  select(c("ls","lsj_g",
           starts_with("Brand_"), 
           starts_with("Country_"),
           "priceur",
           "type", "slr", "elect", "optical","pixeltot")) 
colnames(model4_data)

logit2 <- lm(ls ~  . -Brand_CANON -Country_AUSTRIA , data = model4_data)
summary(logit2)

```

```{r nlogit3}

model5_data <- cameras %>%
  select(c("ls","lsj_g",
           starts_with("Brand_"), 
           starts_with("Country_"),
           "priceur",
           "type", "slr", "elect", "optical","pixeltot",
           starts_with("i"))) 
colnames(model5_data)

ivlogit2 <- ivreg(ls ~ Brand_FUJIFILM + Brand_KODAK + Brand_NIKON + Brand_OLYMPUS + Brand_PANASONIC + Brand_SAMSUNG + Brand_SONY  
                +Country_BELGIUM + Country_BULGARIA + `Country_CZECH REPUBLIC` + Country_DENMARK + Country_FINLAND + Country_FRANCE + Country_GERMANY 
                + `Country_GREAT BRITAIN` + Country_GREECE + Country_HUNGARY + Country_IRELAND + Country_ITALY + Country_NETHERLANDS + Country_POLAND + Country_PORTUGAL 
                + Country_ROMANIA + Country_SLOVAKIA + Country_SLOVENIA + Country_SPAIN + Country_SWEDEN+ priceur + type + slr + elect+optical + pixeltot + (priceur +
                lsj_g) | # need to include again all the regressors in the ivreg function
                Brand_FUJIFILM + Brand_KODAK + Brand_NIKON + Brand_OLYMPUS + Brand_PANASONIC + Brand_SAMSUNG + Brand_SONY 
                + Country_BELGIUM + Country_BULGARIA + `Country_CZECH REPUBLIC` + Country_DENMARK + Country_FINLAND + Country_FRANCE + Country_GERMANY 
                + `Country_GREAT BRITAIN` + Country_GREECE + Country_HUNGARY + Country_IRELAND + Country_ITALY + Country_NETHERLANDS + Country_POLAND + Country_PORTUGAL 
                + Country_ROMANIA + Country_SLOVAKIA + Country_SLOVENIA + Country_SPAIN + Country_SWEDEN + type + slr + elect+optical + pixeltot  
                  #instruments start here 
                 + i1_con + i2_con +  i3_con + i4_con +  i7_con + i1_type + i2_type + i3_type + i4_type + i7_type +   i1_slr +  i2_slr +   i3_slr +  i4_slr +    i7_slr +
                  i1_elect +  i2_elect +  i3_elect +  i4_elect +  i7_elect +  i1_optical +    i2_optical + i3_optical +  i4_optical +  i7_optical +  i1_pixeltot +
                  i2_pixeltot +   i3_pixeltot + i4_pixeltot + i7_pixeltot 
                , data = model5_data) # excluded Brand_CANON + Country_AUSTRIA
summary(ivlogit2)

```


## Price elasticities based on logit

```{r elasticities}
library("broom")

sum <- tidy(ivlogit1)
sum()

alpha = coef(summary(ivlogit1))["priceur","Estimate"]
cameras$own = alpha*cameras$priceur*(1-cameras$share)
cameras$cross = -alpha*cameras$priceur*cameras$share

```

```{r neat}
cameras %>%
  group_by(Country, Brand) %>%
  summarise(mean_own = mean(own)) %>%
  tidyr::pivot_wider(id_cols= Country, names_from = Brand, values_from = mean_own) %>%
  ungroup() %>%
  flextable()

```


```{r elasticities_cross}

cameras %>%
  group_by(Country, Brand) %>%
  summarise(mean_cross = mean(cross)) %>%
  tidyr::pivot_wider(id_cols= Country, names_from = Brand, values_from = mean_cross) %>%
  ungroup() %>%
  flextable()


```

## Merger simulation

```{r antitrust}


```




